---
title: "Probability of Recession"
author: "William Chiu"
date: "2022-11-13"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = TRUE)

```

## Summary

Forecast the probability of a recession in the next 3 months using the following predictors:

1.  Spread between 10Y CMT and Effective Federal Funds Rate

## Extract Historical Data

Refer to this [vignette](https://cran.r-project.org/web/packages/fredr/vignettes/fredr.html) for FRED data access.

```{r}
library(tidyverse)
library(lubridate)
library(caTools)
library(scam)
library(fredr)
library(effects)
library(car)
library(MLmetrics)

series_id <- c("FEDFUNDS", "GS10", "USREC")

full_data <- map_dfr(series_id, function(x) {
  fredr(
    series_id = x,
    observation_start = as.Date("1950-01-01"),
    observation_end = as.Date("2022-12-01")
  )
})
```

## Pivot Wider

```{r}
full_data_wide <- full_data %>% 
  select(date, series_id, value) %>% 
  pivot_wider(id_cols=date, names_from = series_id, values_from = value)
```

## Recession in next 3 months

```{r}
full_data_wide <- full_data_wide %>% 
  arrange(date) %>% 
  mutate(USREC_LEAD1 = lead(USREC, 1),
         USREC_LEAD2 = lead(USREC, 2),
         USREC_LEAD3 = lead(USREC, 3),
         USREC_3MOS = pmax(USREC_LEAD1, USREC_LEAD2, USREC_LEAD3)) %>% 
  drop_na()
```

## Calculate Features/Predictors

```{r}
full_data_wide <- full_data_wide %>% 
  mutate(SPRD_10YCMT_FEDFUNDS = GS10 - FEDFUNDS) %>% 
  select(-USREC, -USREC_LEAD1, -USREC_LEAD2, -USREC_LEAD3)
```

## Split Train/Test

```{r}
set.seed(1)

train_id <- sample.split(full_data_wide$USREC_3MOS, SplitRatio = 0.80)

train_data <- full_data_wide[train_id,]
test_data <- full_data_wide[!train_id,]

summary(train_data)
```

## Logistic Regression

```{r}
logit_mod <- glm(USREC_3MOS ~ SPRD_10YCMT_FEDFUNDS, data=train_data, family=binomial)

summary(logit_mod)
```

## Logit with Knot

```{r}
logit_mod_knot <- glm(USREC_3MOS ~ SPRD_10YCMT_FEDFUNDS + 
                        pmax(0,SPRD_10YCMT_FEDFUNDS),
                      data=train_data, family=binomial)

summary(logit_mod_knot)
```


## Shape-Constrained GAM

```{r}
scam_mod <- scam(USREC_3MOS ~ s(SPRD_10YCMT_FEDFUNDS, bs="mpd"),
                 data=train_data, family=binomial())

summary(scam_mod)

plot(scam_mod,pages=1,shade=TRUE)

```

## Effect Plot

```{r}
plot(predictorEffects(logit_mod),
     axes = list(
       grid = TRUE,
       x = list(rug = FALSE),
       y = list(type = "response")
     ))
```

## Effect Plot with knot

```{r}
plot(predictorEffects(logit_mod_knot),
     axes = list(
       grid = TRUE,
       x = list(rug = FALSE),
       y = list(type = "response")
     ))
```

## Null Model

```{r}
null_mod <- glm(USREC_3MOS ~ 1, data=train_data, family=binomial)

summary(null_mod)
```

## Performance Metric

```{r}
test_preds <- predict(logit_mod, newdata=test_data, type="response")
null_preds <- predict(null_mod, newdata=test_data, type="response")
knot_preds <- predict(logit_mod_knot, newdata=test_data, type="response")


perf <- function(lst_preds, f_metric=caTools::colAUC, metricname="ROC-AUC"){
  map_dfr(lst_preds, function(x){
  f_metric(x, test_data$USREC_3MOS)
}) %>% 
  pivot_longer(everything(), names_to="model", values_to=metricname) %>% 
  knitr::kable()
}

myPreds <- list(logit_reg=test_preds, null_model=null_preds,
                knot_reg=knot_preds)

perf(myPreds, caTools::colAUC, "ROC-AUC")
perf(myPreds, MLmetrics::LogLoss, "LogLoss")
```

## Probability of Recession (11/11/2022)

10Y CMT = 3.82 

Eff. FFR = 3.84

```{r}
curr_data = data.frame(SPRD_10YCMT_FEDFUNDS = 3.82 - 3.84)

mods <- list(logistic_reg=logit_mod,
                              scam_mod=scam_mod,
                              knot_mod=logit_mod_knot,
                              baseline=null_mod)

map_dfc(mods, function(x){
  predict(x, newdata=curr_data, type="response")
}) %>% 
  pivot_longer(everything(), names_to = "model",
               values_to = "prob_rec") %>% 
  knitr::kable(.)

```
Relative to the historical baseline, a slightly negative spread between the 10Y CMT and effective federal funds rate increases the chances of a recession in the next 3 months.